---
title: "Public Health Graphical Display Using R"
output:
  html_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

## Link to Folder with HLML version of this document and all files for this workshop:
## https://goo.gl/DCVQCk


# Chapter 0.1 --- Data Sets --- 

```{r warning=FALSE }
myPlace <- getwd()
source('dataSetsLocal.R')
```

```{r eval=F, warning=FALSE, message=F}
library(gsheet)
source('dataSetsWeb.R')
```

```{r include=FALSE, warning=FALSE, echo=FALSE}
library(classInt)
library(maptools)
library(RColorBrewer)
library(epitools)
library(rmarkdown)
library(RColorBrewer)
library(shiny)
library(dplyr)
```


### 10/2018 Draft - Work in Progress

Michael C. Samuel, Dr.P.H. <br>
Data Scientist / Senior Epidemiologist <br>
California Department of Public Health <br>
Michael.Samuel@cdph.ca.gov <br>
925-285-2926 

Please contact Michael Samuel to report errors or for other suggestions 

# Chapter 0.0 --- Introduction ---

R (like its parent S) is an unparalleled tool for producing high-quality visual displays. The power of R for visual display is based on its complete integration of data and graphics objects. Figures are produced using a wide range of excellent high and low level built-in graphical tools, project-specific custom code including functions, and a phenominal array of add-on "packages" built by users all over the world.

This overview describes the basic operation of the R graphical procedures, teaching by example, using data and examples from public heath surveillance. The details covered here only begin to scratch the surface of the power and complexity of R's graphical display functionality.

You can learn a lot by looking at the built-in R examples and demos, with all code:

demo(graphics)  
example(barplot)  
example(title)   
example(hist)  
example(plot)  
example(matplot)  
example(boxplot)  
demo(plotmath)

par(ask=F)

Help/Getting Started
Read/review "Graphical Procedures" Chapter (Chapter 12) in "An Introduction to R"

"Quick-R" site with "cookbook" approach to using R including graphics
http://www.statmethods.net/index.html

Data Analysis and Graphics Using R - An Example-Based Approach
http://maths-people.anu.edu.au/~johnm/r-book/daagur3.html


LOTs of helpful info on R graphics at:
http://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html (including .pdf of introductory chapter and chapters on the grid and lattice packages and at: http://www.stat.auckland.ac.nz/~paul/

Gallary of R graphics and code, and some related links at: http://addictedtor.free.fr/graphiques/

Data Analysis and Graphics Using R - An Example-Based Approach. John Maindonald and John Braun. 2nd edn, Cambridge University Press, January 2007. Related info:
http://maths.anu.edu.au/~johnm/r-book/daagur3.html

R Graphics-Related Packages and Other Links and Resources
In addition to the core graphically functionality of R, the "lattice" and "grid" packages are included with the main R system download.  They add a wide range of tools, particularity for complex multi-row and or muti-column figures,and for combining types of charts (e.g. Bar and line charts).  When using lattice/grid graphics, you do not/can not use the "regular" R graphics functions or parameters.

"Trellis Graphics is a framework for data visualization developed at the Bell Labs by Rick Becker, Bill Cleveland et al, extending ideas presented in Bill Cleveland's 1993 book _Visualizing Data_. Lattice is best thought of as an implementation of Trellis Graphics for R...Lattice is built upon the Grid Graphics engine for R ... and requires the 'grid' add-on package. Type 'help(package = lattice)' to see a list of (public) Lattice graphics functions for which further documentation is available".

"ggplot2 is a plotting system for R, based on the grammar of graphics, which tries to take the good parts of base and lattice graphics and avoid bad parts. It takes care of many of the fiddly details that make plotting a hassle (like drawing legends) as well as providing a powerful model of graphics that makes it easy to produce complex multi-layered graphics."
Find out more at http://had.co.nz/ggplot2, and check out the over 500 examples of ggplot in use"

There are hundreds of other graphics-related packages that may be of use generally or for specific types of plots. 

Other Resources:  
Parameters and functions cheat-sheet: https://cran.r-project.org/doc/contrib/Baggott-refcard-v2.pdf  
Color advice: http://colorbrewer2.org/  
R color cheatsheet: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf  
R Studio cheat-sheets: https://www.rstudio.com/resources/cheatsheets/  
Anything Zev Ross says: http://zevross.com/blog/  

# VISUAL DISPLAY IN BASE R

# Chapter 1 --- Graphical Parameters ---

```{r} 
barplot(1:10)
```  

R plotting commands are divided into two main groups, High-Level, that are generally used to make new plots, and Low-Level, that are generally used to add things to plots.  BOTH use *graphical parameters* to manipulate and customize plots.  

Global parameters: change in all cases
```{r warning=FALSE} 
parsave <- par()       # save the default parameters
``` 
 
```{r} 
par(las=1)  # specify horizontal axis labels label
barplot(1:10)
```  


```{r warning=FALSE} 
par <-par(parsave)     # return to default parameters
```  

Local parameters: change in the specific figure
```{r}
barplot(1:10, col="blue")
```	

List of all parameters
```{r eval=FALSE} 
help(par)
```

A large proportion of what you need to know about the parameters is on the R Reference card noted above. 
Parameters and functions cheat-sheet: https://cran.r-project.org/doc/contrib/Baggott-refcard-v2.pdf  

# Chapter 2 --- Simple Frequency Distribution Plots (e.g. histograms, boxplots) ---

There are *many* excellent R graphical routines for looking at the frequency or distribution of a variable, including histograms, box plots, "strip plots", and "density plots".

hist(rnorm(100))
hist(rnorm(10000))
```{r}
hist(rnorm(10000),breaks=50)
```

### Get to know the gonorrhea male rate variable before plotting

```{r}
attach(usSTD) # Attach the usSTD data frame for convenience
gcm
sort(gcm)
summary(gcm)

hist(gcm,breaks=10)	 # Histogram 
boxplot(gcm)			   # Box-plot 
plot(density(gcm))	 # Density plot

detach(usSTD)
```

# Chapter 3 --- Simple Summary Plots  (e.g. bar charts, pie charts) ---

R has countless great routines to display summary data, like counts, percents and rates-- bar plots, pie charts and dot charts are all such plots


## Simple Vertical Barplot 

```{r}
attach(ca.gc)	# Attach the ca.gc data frame for convenience
```

```{r eval=F}
barplot(m.rate )
```
```{r eval=F}
barplot(m.rate,names.arg=names.short)
```
```{r}
barplot(m.rate,names.arg=names.short,col=4)	# bar chart with names and color
```

##  Horizontal Bar Chart
```{r warning=FALSE }
parsave <- par()         
par(mar=c(4,15,2,2))     # specify margins for long labels
barplot(m.rate,names.arg=as.character(race.ethnic),col=4,horiz=TRUE,las=1,axis.lty=1)
par(parsave)             
```

## Cluster Bar Chart
```{r}
temp <- barplot(rbind(m.rate,f.rate),names.arg=names.short, col=c(2,4),horiz=TRUE,
        las=1,   
        beside=TRUE,
        axis.lty=1)          
legend(x=125,y=5,legend=c("Male","Female"),fil=c(2,4))
```

## Stacked Bar Chart
```{r}
temp <- barplot(rbind(m.cases,f.cases),names.arg=names.short,col=c(2,4), horiz=TRUE,las=1, 
		    beside=FALSE)     #Stacked Bar Chart         
legend(x=3000,y=2,legend=c("Male","Female"),fil=c(2,4))
```

parameter notes:
### horiz         - for horizonal bars
### beside=TRUE   - for cluster bar chart
### beside=FALSE  - for stacked bar chart
### las           - changes the orientation of labels relative to the axis
### axis.lty      - is a parameter that controls the axis line

## Pie Charts
```{r}
pie(m.cases,labels=names.short)
```
## Dot Charts
```{r eval=F}
dotchart(m.cases,labels=as.character(race.ethnic),cex=.7)   #CEX is a parameter to change relative font size
```
```{r}
dotchart(cbind(m.cases,f.cases),labels=as.character(race.ethnic),cex=.7)
```

detach(ca.gc)


# Chapter 4 --- Plots Examining Relationships ---

R is particularly powerful for looking at relationships in data, and has many related types of graphical routines for this area. A common function that is very powerful and flexible for looking at relationships is the "plot" function. Two simple applications of the plot function are a line chart, showing a trend over time, and a scatter plot showing the relationship between two continuous, or basically continuous, variables.

### Generate some vectors for plotting

```{r}
x0 <- 1:50
x  <- x0 + rnorm(50)*5
y  <- x  + rnorm(50)*5
```

## Line Graphs and Scatterplot

```{r eval=F}
plot(x )               #basic ?
```

```{r eval=F}
plot(x0,x)            #basic plot
```

```{r}
plot(x0,x,type="l")   #basic line gr
```

##  line/point "type" parameter
```{r}
plot(x0,x,type="b")
```

```{r}
plot(x,y)	            #basic scatter plot
```

```{r eval=0}
plot(x,y,type="l")    #basically doesn't make sense
```


```{r}
attach(gc.trend) # for convenience
plot(Year,California,type="l")
lines(Year,US,col=4)           	# Add line to create mult-line graph
```

Use matplot to create multi-line graph with one command; I generally don't use 'for real' becuase I generally want more control, but good here to show 'line types' 			

```{r}
matplot(1:10,matrix(1:100,ncol=10),type="l")
```

### Line Types 
http://www.statmethods.net/advgraphs/parameters.html
http://www.stat.tamu.edu/~jkim/Rlinestyle.pdf
http://cran.r-project.org/doc/contrib/Short-refcard.pdf


## Boxplot by Categorical Variable

SFMHS Data

```{r eval=F}
attach(sfmhs) # for convenience
plot(SEXPREF,CD4) # a scatterplot...
```

```{r eval=F}
boxplot(SEXPREF,CD4)
```

```{r eval=F}
plot(as.factor(SEXPREF),CD4) # same as boxplot...
```

```{r}
attach(sfmhs) # for convenience
CD4[CD4 > 2000] <- NA
group <- SEXPREF-1
group[group==1 & ARV==1] <- 2
group <- factor(group,labels=c("Het","G/B ARV-","G/B ARV+"))
plot(group,CD4,varwidth=TRUE,notch=TRUE)
detach(sfmhs)
```

## Plotting symbols proportional to data
```{r}
region      <- c(1,1,1,2,2,2,3,3,3)
prevalence  <- c(2,3,2,3,5,4,7,4,8)  
number      <- c(3,4,3,2,6,5,8,4,7)
symbols(region,prevalence,circles=number,inches=.5)
```

## "Ploting" 2 by 2 tables (SKIP)

```{r}
outbreak <- matrix(c(200,40,5,23),byrow=TRUE,nrow=2,dimnames=list(c("Ate It","Didn't"),c("Ill","Well")))
outbreak
chisq.test(outbreak,correct = F)
library(Epi) # Epi Package
twoby2(outbreak)
```


```{r}
mosaicplot(outbreak,col = "blue",xlab = "Ill",ylab = "Exposed")
```

```{r}
assocplot(outbreak)
```

```{r}
fourfoldplot(outbreak)
```

LOTS and LOTS of other types of plots in R


# Chapter 5 --- NUTS AND BOLTS ---

## Titles and Axes Titles
Titles can be created using the "title" function, the "main" parameter in a plot function, or the "mtext" function discussed later

x and y axis labels can be created using the "xlab" and "ylab" parameters in a plot function, or using the "mtext" function 

### main, xlab, & ylab approach:

```{r eval=F}
attach(usSTD) # Back to the US STD Data
plot(gcf,gcm)
```

```{r}
attach(usSTD) # Back to the US STD Data
plot(gcf,gcm,main="US Male versus Female Gonorrhea Rates by State",
             xlab="Female Rate (per 100,000)",
             ylab="Male Rate (per 100,000)")
```

## Adding text to axes using mtext and axis:
mtext is a powerful function to any type of text (including numbers) to a specific margin of a figure, at specific coordinates, on a specific "line"; uses the coordinates generated by the plot function

axis is also powerful for adding text to the margins (it can also be used to adjust the tics and other aspects of the margins)

### mtext

```{r}
x <- barplot(1:5,col=1)
my.labels <-c("First","second","Third","Fourth","Fifth")
my.marks <- c(1,3,5)
mtext(my.labels[my.marks],side=1,at=x[my.marks])
```

"at" specifies where along the axis to place the labels; you WILL use this...

### using the axis function

```{r}
barplot(1:5,col=1)
axis(1, at=x[my.marks],labels=my.labels[my.marks],cex.axis=.5)
axis(3, at=x[my.marks],labels=my.labels[my.marks],cex.axis=.5)
```

## Legends and Labels

```{r}
attach(gc.trend)
plot(Year,California,type="l")
lines(Year,US,col=4) # add US line 

# Add a legend using STANDARD LEGEND function arguments

legend(1920,500,legend=c("California","United States"),cex=.8,lty=1,col=c(1,4),text.col=c(1,4))
```


## Adding text to the body of a figure using the Text Function

TEXT is a useful function for adding text strings to any specified location inside a figure based on x,y coordinates

add some annotation using text function, specifying coordinates and using locator

text(1985,580,"Peak")
text(locator(),"Nadir")

The TEXT function can be used to directly label lines (better than using a legend in many situations) or to build complex legends, when the legend function is inadequate

```{r}
plot(Year,California,type="l",xlim=c(1920,2015),
     main="Gonorrhea Trend, CA and US, Through 2002",
     xlab="Year",
     ylab="Rate (per 100,000)")
lines(Year,US,col=4) # add US line

text(2003,California[length(California)],"California",cex=.7,adj=c(0,.5))
text(2003,US[length(California)],"United States",cex=.7,adj=c(0,.5),col=4)

detach(gc.trend)
```


## Text "Data" 
### Using text for detailed labeling marking of points

Read in file that has "fips" codes and state abbreviations and merge the usSTD data set and this fips data set

```{r}
usSTD   <- read.table(paste0(myPlace,"/RawData/ctgcps2002.csv"),sep=",",na.strings="",header=T)
fips    <- read.table(paste0(myPlace,"/RawData/fipsdata.csv"),sep=",",na.strings="",header=T,as.is=T)
usWork  <- merge(usSTD,fips,by.x="FIPSCODE",by.y="FIPScode")
attach(usWork)

plot(gcm,ctm,
     pch=1,cex=1.75,col=4,las=1,
     xlim=c(0,max(gcm,gcf)),
     ylim=c(0,max(ctm,ctf)),
     xlab="GC Rate",
     ylab="CT Rate")
text(gcm,ctm,StateAb,cex=.4,col=4)
points(gcf,ctf,pch=1,cex=1.75,col=2)
text(gcf,ctf,StateAb,cex=.4,col=2)
legend(400,600,c("Female","Male"),col=c(2,4),pch=1)

```

### Can use "text" function to make custom legend
text(450,600,"Female",col=2,adj=0,cex=.75)
text(450,550,"Male",col=4,adj=0,cex=.75)
points(440,600,pch=1,col=2,cex=1.75)
points(440,550,pch=1,col=4,cex=1.75)


### Line type (lty) and line width (lwd) reference plot
```{r}
matplot(1:10,matrix(rep(1:10,10),nrow=10,byrow=TRUE),type="l",col="black",lty=1:10,lwd=1:10)

```
or one can use the character strings "blank","solid","dashed","dotted","dotdash","longdash", "twodash", or go crazy and a string of up to 8 characters (from `c(0:9,"A":"F")') may be given, giving the length of line segments which are alternatively drawn and skipped.  

### Plotting symbol reference
```{r}
plot(rep(c(1,2,3),each=10),rep(0:9,3),pch=0:29,xlim=c(.5,3.5),xaxt="n",yaxt="n",ylab=NA,xlab=NA)
title("pch Plotting Symbols")
text(rep(c(1,2,3)-.2,each=10),rep(0:9,3),as.character(0:29),cex=.7)
```

### Plotting symbol line type reference 
```{r}
matplot(1:10,matrix(rep(1:10,10),nrow=10,byrow=TRUE),type="o",pch=1:10,lty=1:10,col="black")

```

## Ticks and Grids

### Ticks
```{r}
attach(usSTD)
```

```{r}
plot(gcf,gcm)
plot(gcf,gcm,tck=-0.04)
axis(side=1,at=(1:40)*10,labels=FALSE)

```

### Rugs! (tics at data values)
R has a very nice rug() function to add tick marks at each occurrence of data point; 
```{r}
plot(gcf,gcm)
rug(gcf)
rug(gcm,side=2)
detach(usSTD)
```
## Grid Lines
Grid lines can be drawn in using axis(), grid(), lines(), or with the abline() functions, and in other ways. All of these have some strengths and some weaknesses; in any case grid lines should usually be in light colors, and often dashed, so that they are not too prominent. abline is usually my choice.

### using axis
```{r eval=F}
attach(gc.trend)
plot(Year,California,type="l",main="California GC Trend 1913-2002",xlab="Year",ylab="Rate (per 100,000)")
axis(2,tck=1,lty="dashed",lwd=0.2,col="gray")
```

### using grid
```{r eval=F}
plot(Year,California,type="l")
grid()
```
grid(nx=NULL,ny=21)
grid(nx=NA,ny=21)

### using abline
```{r}
plot(Year,California,type="l")
abline(h=c(0:12)*50,lty=2,col="gray")
```



# Scale

## 2 Axis Chart 
### (here with bars and lines, but the key for anything like this is changing the "left" Y-asis coordiantes 


California Syphilis Surveillance data; Early MSM cases by HIV status by half year interval

Some review of some R basics.....


```{r}
ca.msm.sy <- cbind(c( 6, 7,65,47,126,188,343,463,465,394),
                   c( 0, 2,42,57, 67,101,161,227,291,206),
                   c(31,28,22,12, 29, 47, 76, 74, 61, 67))
dimnames(ca.msm.sy) <- list(period=paste(rep(1999:2003, each = 2),c("a","b"),sep=""),status=c("HIV+","HIV-","HIV?"))
           
all        <- apply(ca.msm.sy,1,sum)
all.w.data <- apply(ca.msm.sy[,1:2],1,sum)
per.hiv    <- 100*ca.msm.sy[,"HIV+"]/all.w.data
nn         <- length(all)

parsave <- par()
par(mar=c(5,4,4,4)+.1)

junk <- barplot(all,las=2,col="blue")
mtext(side=2,line=3,"Number of Cases",col="blue")

#  this is the MAGIC LINE; use existing X-coordiantes, and change Y-coordiantes to 0 to 100.
par(usr=c(par()$usr[1:2],0,100))  

lines(junk,per.hiv,lwd=2,col="red")
axis(side=4,las=2)
mtext(side=4,line=2.5,"Percent HIV+",col="red")

legend(3,100,legend=c("N","%HIV+"),lty=c(NA,1),lwd=3,fil=c(4,NA),col=c("blue","red"),bty="o",border="white",adj=-0,cex=1.3)

```


Another approach to a plot with separate y1 and y2 axes, from www.demog.berkeley.edu/faq/node21.html. This will create a scatter plot displaying two sets of points, one corresponding to the y axis on the left and one to the y axis on the right

```{r}
plot(x<-rnorm(100),y<-rnorm(100))
z<-rnorm(100)*250
par(new=T)   ## Tell R not to reinitialize graphic device for subsequent plots
plot(x,z,col='blue',axes=F)
axis(side=4,col.axis='blue')
par(new=F)
```


## Banking to 45?
```{r}
nn <- 50
plot((1:nn)*.92,sin((1:nn)*.92))
```

DRAG y-axis together!!!

## Smoothing
LOTs of good smoothing functions in R including lowess function 
```{r}
attach(gc.trend)
plot(Year,California,type="l",main="California Gonorrhea Rates by Year")
lines(lowess(Year,California),lty=2)
lines(lowess(Year,California,f=1/3),lty=2)
detach(gc.trend)
```
# Chapter 6 --- SMALL MULTIPLES ------------------------- 

The mfrow=c(n,k) parameter is the fundamental parameter to manually set up "small multiple" graphics; this parameter sets the graphic area to display n row by k columns of figures per page; mfcol is the same but fills by column.

```{r}
par(mfrow=c(3,2),mar=c(0,0,0,0)) # sets page to 3 rows of 2 columns
attach(usSTD)
plot(gcm,ctm)
plot(gcf,ctf)
plot(gcm,sym)
plot(gcf,syf)
plot(ctm,sym)
plot(ctf,syf)
par(mfrow=c(1,1))
```


### Layout

 'layout' divides the device up into as many rows and columns as
     there are in matrix 'mat', with the column-widths and the
     row-heights specified in the respective arguments.
 Usage:
     layout(mat, widths = rep(1, ncol(mat)),
            heights = rep(1, nrow(mat)), respect = FALSE)


```{r}
layout(matrix(c(1,1,1,2,3,3,2,4,5),3,3,byrow=TRUE))

layout.show(5)

barplot(1:3,main='Top row, all columns')
barplot(1:10,main='2nd AND 3rd rows, 1st column')
plot(1:10,1:10,main='2nd  row, 2nd and 3rd columns')
pie(1:5,main='3rd row, 2nd column')
pie(5:1,main='3rd row, 3rd column')

layout(1)

```





There are also a number of automated routines in R for creating small multiple-type displays

### pairs
Pairs() is one such function that creates a series of all pair-wise scatter plots based on columns on the input matrix


```{r}
usSTDsmall <- usSTD[,c(13,14,15,18,21)]
pairs(usSTDsmall) # same as "plot(usSTDsmall)"

```


There are also entire other packages, largely for the display of small multiple-like figures; these include the grid and lattice packages 

A (the?) basic function for display small multiples based on stratification variables is the xyplot() function, and its related functions 


```{r}
attach(sfmhs)
group <- SEXPREF-1
group[group==1 & ARV==1] <- 2
group <- factor(group,labels=c("Het","G/B HIV-","G/B HIV+"))
library(lattice)
xyplot(HEIGHT~WEIGHT|group)
```

### Amazing displays can be created with these "small multiple-oriented packages!=

```{r eval=F}
par(ask=TRUE)
library(grid)
demo(lattice)
example(xyplot)   # show this one

example(coplot)
par(ask=FALSE)
```


# Chapter 6 --- Color in R  ----------------------------------------------------------------------------------------- 

Making and using colors is R is extraordinarily flexible.  It is easy to choose, make, name, and change colors for any componet of a chart. There are LOTS of "built-in" colors and built-in color ramps.  There are lots of great resources for learning more about colors in R, inluding this: http://research.stowers.org/mcm/efg/Report/UsingColorInR.pdf

The colors() function lists the names of the 657 colors R has stored internally 
```{r eval=FALSE}
palette("default")
colors()
```


### color by NUMBERS!
```{r echo=FALSE}
source(paste0(myPlace,"/Color Resources/colorSquareNumber.R"))
```

### color by NAMES!
```{r echo=FALSE}

source(paste0(myPlace,"/Color Resources/colorSquareNAME.R"))
```


```{r}
#some on-line resources for color numbers and names
#{r echo=FALSE} 
#system2('open', args = "https://rstudio-pubs-static.s3.amazonaws.com/3486_79191ad32cf74955b4502b8530aad627.html") -->
#system2('open', args = "http://www.bxhorn.com/Downloads/RColors1_bxhorn.pdf") -->
#system2('open', args = "http://research.stowers.org/mcm/efg/R/Color/Chart/ColorChart.pdf") -->

```




There are 6 base R functions for building special continuous colors or "color ramps"--rainbow, heat.colors, topo.colors, terrain.colors, cm.colors (cyan-magenta), and grey. The main argument to these is n, the number, of equally spaced colors; the rainbow function has other arguments to further control the exact colors; arguments to grey (or gray) function are between 0 and 1, and indicate degree of "grayness" from black to white. 


```{r}
# {r eval=FALSE} for MS Word verion...
par(mfrow=c(3,2),oma = c(0, 0, 3, 0))
barplot(1:100,col=rainbow(100),main="rainbow")
barplot(1:20,col=heat.colors(20),main="heat.colors")
barplot(1:10,col=topo.colors(10),main="topo.colors")
barplot(1:10,col=terrain.colors(10),main="terrain.colors")
barplot(1:11,col=grey(0:10/10),main="grey")
barplot(1:10,col=cm.colors(10),main="cyan-magenta")
mtext(outer = TRUE,"base color functions",side=3)
par(mfrow=c(1,1))
```



There are other functions to create color "ramps". One widely used is from the "RColorBrewer" package, and based on the fantatic webiste http://colorbrewer2.org. The package provides great options for sequential, Diverging, and Qualtiative color pallettes, based on research.

sequential palettes (3-9 values): 
Blues BuGn BuPu GnBu Greens Greys Oranges OrRd PuBu PuBuGn PuRd Purples RdPu Reds YlGn YlGnBu YlOrBr YlOrRd
diverging palettes (3-11 values): 
BrBG PiYG PRGn PuOr RdBu RdGy RdYlBu RdYlGn Spectral


```{r}
# {r eval=FALSE} for MS Word verion...
par(mfrow=c(3,2),oma = c(0, 0, 3, 0))
display.brewer.pal(9,"Blues");    barplot(1:3,col= brewer.pal(3,"Blues"))
display.brewer.pal(9,"YlOrBr");   barplot(1:9,col= brewer.pal(9,"YlOrBr"))
display.brewer.pal(11,"RdBu");    barplot(1:5,col= brewer.pal(5,"RdBu"))
mtext(outer = TRUE,"Color Brewer Color",side=3)
par(mfrow=c(1,1))

```


The "colorRamp" and "colorRampPalette" functions create color ramps too.


Anything not built-in can be made, using any of the color systems, including the "rgb" (red, green, blue) or "hsv" (hue, saturation, value) functions,

Here we use "rgb" to create a vector, myColor, of five new colors with names based on a red, a green and a blue vector. And we use "hsv" to create a sigle color, myBlueToo, using hue, saturation, and value values.


```{r}
myColor <- rgb(c(255,255,240,255,0),c(127,20,230,192,255),c(80,147,140,203,127),
         names=c("m.o1","m.p1","m.y1","m.p2","m.g1"),maxColorValue=255)

myBlueToo <- hsv(.5,.5,.5)

```

For bars and lines in typical figures, the "col" argument controls the bar or line colors; it can be one argument so the color is the same for all bars and lines, or, using a vector, one can specify the color for each bar or line 

```{r}
par(mfrow=c(3,2))
barplot(1:6)                                      # uses (single) default color
barplot(1:20,col="thistle2")                      # uses single predefined color
barplot(1:20,col=myBlueToo)                       # uses single newly made color
barplot(1:11,col=0:10,names=c(0:10),cex.names=.6) # uses set of colors from default palette
barplot(1:5,col=myColor)                          # uses set of newly created colors
barplot(1:11,col=myColor)                         # uses set of newly created colors
par(mfrow=c(1,1))
```


## Palette Function 
The palette() function displays the current default color pallet; by default is uses the color names if they exist, otherwise it will use the hexadecimal representation of the color; the palette function can also be used to CHANGE the current palette.

Not sure about this: "The color palette and referring to colors by number (see e.g. par) was provided for compatibility with S: in R it is almost always better to specify colours by name."


```{r}
palette()
```


```{r}
par(mfrow=c(2,1))
barplot(1:5,col=1:5)
barplot(1:5,col=myColor)
par(mfrow=c(1,1))
```


```{r}
par(mfrow=c(2,1))
palette(myColor) # makes "mycolors" the default colors
barplot(1:5,col=1:5)
barplot(1:5,col=myColor)
palette("default")  # and return to the default palette
par(mfrow=c(1,1))
```



## Col2rgb function 
Col2rgb function presents shows the the red, green and blue values for a given vector of color names, hexadecimal valeus, or numeric value in the current color palette.

```{r}
col2rgb("white")
col2rgb(3)            # 3 here means palette()[3]
```




A little example of making a color set, using it for the default palette, and using it for all colors in a figure. This type of system is particularly useful for figures that are created regularly, and where, for example, the background color may need to change from white to dark blue on occasion.


```{r}
# create two color vectors
colset1 <- c("blue","black","red","white")
colset2 <- c("yellow","white","cyan","blue")

# and specifiy colors in code with NUMBERS, not names:

#parsave <- par()  # and reset palate
#par(bg=4)

plot1 <- function() {
  t.x <- barplot(c(1,2,3,8),col=1,col.axis=1,main="Really Big Deal",col.main=2,bg=4)
  axis(side=1,at=t.x,labels=c("a bit","some","more","lots"),las=2,col.axis=2,col=2)
  axis(side=2,col=1,labels=F)
  mtext(side=2,line=3,"One Thing",col=1)
  par(usr=c(par()$usr[1:2],0,100))
  lines(t.x,c(10,20,60,20),lwd=2,col=3)
  axis(side=4,las=2,col.axis=3,col=3)
  mtext(side=4,line=2.5,"Another",col=3)  }

palette(colset1)   # make this vector the default palette
plot1()
palette("default")
```

```{r}
palette(colset2)   # now re-run with the other color vector
plot1()
palette("default")
```


## Transparent  Color
R has a nice system for creating transparent color, to enable seeing objects behind other objects in an R figure.  It is important to test the results of using transparent color on different devices (e.g. the console, send to a PDF file, and printed) since the appearance of the transparent color can differ substantially.  The Alpha argument to rgb color construction function determines degree of transparency of a color


```{r}
col1 <-"blue1"
col2 <-"red1"
col2 <- rgb(255,0,0,alpha=200,max=255)

t.x  <- 1:5
t.y1 <- c(1,1,3,3,5);       t.s1 <- c(1,5,3,3,3)
t.y2 <- c(1.3,1.5,3,3,4.9); t.s2 <- c(1,2,5,3,3)  

symbols(t.x,t.y1,circles=t.s1,inches=.5,bg=col1)
symbols(t.x,t.y2,circles=t.s2,inches=.5,bg=col2,add=TRUE)
```

## Shading / Fill
R seems to have limited options, except for shades of grey, for black and white fill; the one option is has is for the density and angle of diagonal lines, a fill much criticized by Tufte and others
```{r}
barplot(1:5,angle=c(10,30,45,60,90),density=c(3,5,5,10,100),col="blue")
```





# Chapter 7 --- SAVING AND PRINTING  ----------------------------------------------------------------------------------------- 


Figures can be copied from the console window and pasted directly into any typical application program (e.g. OpenOffice document or presentation, Word document or PowerPoint). Figures can be copied in many formats using the "Export" button in the output window.

```{r}
barplot(1:3)
```

Figures can also be saved using a number of devise options or after a figure has been made using the saveplot function

Specify output file THEN make plot using graphics "devices/function"(bmp, png, wmf, and others); figure height and width, quality, and point size can be adjusted with parameters for some of these devices; through limited experimentation, I have found that the wmf format seems to work best for projects where high qualtiy images are needed  

```{r} 
jpeg(filename=paste0(myPlace,"/inOut/junkplot1.png"),quality=100,width=1000,height=800)
plot(x<-1:100,x^2)
dev.off()
```

Make plot THEN direct output using savePlot
Fix this some day?
plot(x,x^3)
savePlot(filename=paste0(myPlace,"/inOut/junkplot2.jpg"),type="jpg")
dev.off()


Series of "small multiple" figures using mfrow and multiple pages, indexing across rows (local health jurisdictions) of data frame  

Demonstrates automated method for saving figures to files


```{r}
win.metafile(filename= paste0(myPlace,"/G.out/GCplot%03d.wmf"))  # Integer formatting in file names causes files to be sequentially sent to numbered files

attach(CA.gc.race)
par(mfrow=c(3,5),mar=c(0,1,1,0)+1)

for (i in 1:62) {
t<- barplot(c(black.m.r[i],   black.f.r[i],
              white.m.r[i],   white.f.r[i],
              hispanic.m.r[i],hispanic.f.r[i],
              asian.m.r[i],   asian.f.r[i],
              total.m.r[i],   total.f.r[i]),
         main=region[i],
         col.main="blue2",
         pch=rep(1:5,5),
	       col=rep(1:2,2,each=2))}

dev.off()

detach(CA.gc.race)
```



# Chapter 8 --- Epi Curve ----------------------------------------------------------------------------------------- 

# Epi Curve
## from the MedEpi himself, Dr. Tomas Aragon, and his EpiTools package
http://www.medepi.net/epitools/examples.html

```{r}
library(epitools)

butte <- read.table(paste0(myPlace,"/RawData/buttegc.csv"),sep=',',na.strings='',header=T)
attach(butte)

bdates <- as.Date(DXDATE,format ='%m/%d/%Y')
goodvalues <- !is.na(bdates) & bdates > as.Date('2000-1-1')
bdates.1 <- bdates[goodvalues]
barea.1  <- CITY[goodvalues]
brace.1  <- RACE[goodvalues]

work <- epicurve.months(bdates.1,axisnames=FALSE,segments=TRUE)
axis(1,at=work$xvals,labels=work$cmonth,las=2,cex.axis=.3)
axis(1,at=work$xvals,labels=work$cyear,line=1,cex.axis=.7,tick=FALSE)

work <- epicurve.months(bdates.1,axisnames=FALSE,strata=brace.1, col=1:length(table(brace.1)), min=as.Date("2002-01-01"),max=as.Date('2005-12-31'),legend.text=TRUE)

```



# Chapter 9 --- MAPPING /GIS ----------------------------------------------------------------------------------------- 

### reading the data and shape files
```{r}

iData    <- read.csv(paste0(myPlace,"/RawData/CID-Cases-by-Disease-County-Year-Sex.csv"))
ratedat  <- iData[iData$Disease == "Gonorrhea" & iData$Sex == "Total" & iData$Year == 2014 &  iData$County != "California", c("County","Rate")]

shape    <- readShapePoly(paste0(myPlace,"/shapeFiles/CaliforniaCounty"))
map.dat  <- merge(shape,ratedat, by.x="NAME",by.y="County")
```

### plot with no arguments -- just get the outline
```{r}
plot(map.dat)
```

### thematic map
```{r}
colors1 <- rev(brewer.pal(5, "RdBu"))
brks<-classIntervals(ratedat$Rate, n=5, style="pretty")$brks

plot(map.dat, col=colors1[findInterval(ratedat$Rate, brks,all.inside=TRUE)], axes=F)
legend(x=-118, y=42, legend=leglabs(brks), fill=colors1, bty="n", x.intersp = 1, y.intersp = 1,cex=.8)
title("Gonorrhea Rates by California County, 2014")
```

### SAME map with quantile breaks
```{r}
brks<-classIntervals(ratedat$Rate, n=5, style="quantile")$brks
plot(map.dat, col=colors1[findInterval(ratedat$Rate, brks,all.inside=TRUE)], axes=F)
legend(x=-118, y=42, legend=leglabs(brks), fill=colors1, bty="n", x.intersp = 1, y.intersp = 1,cex=.8)
```

### SAME map with custom breaks
```{r}
rates <- ratedat[ratedat$Rate !=0,"Rate"]
brk0   <- classIntervals(rates, n=4, style="quantile")
brk1   <- round(brk0$brks,1-trunc(log10(median(rates))))
brks   <- c(0,brk1)
plot(map.dat, col=colors1[findInterval(ratedat$Rate, brks,all.inside=TRUE)], axes=F)
legend(x=-118, y=42, legend=c(0,leglabs(brk1)), fill=colors1, bty="n", x.intersp = 1, y.intersp = 1)
```

# Chapter 10 --- Shiny, barely an introduction ----------------------------------------------------------------------------------------- 

```{r eval=F}

# Define the UI
ui <- bootstrapPage(
  numericInput('n', 'Number of obs', min=0,max=1000000,value=50),
  numericInput('bins', 'Number of bins', min=5,max=1000,value=15),
  plotOutput('plotMy')   )

# Define the server code
server <- function(input, output) {
  output$plotMy <- renderPlot({hist(rnorm(input$n),breaks=input$bins,col="cyan")})    }

# Return a Shiny app object
shinyApp(ui = ui, server = server)
```


### Shiny and GIS training exaple
  ".../GIS and Shiny"

### Production apps
  
  ".../0.CBD/myCBD"   (& https://phdataviz.shinyapps.io/myCBD/)
  
   https://pdop.shinyapps.io/ODdash_v1/


# GGPLOT2


dev.off()   ## start with a new graphics device


